Better fusion for filters

### 2014-04-07 2014betterfusionforfilters/section/01-Introduction.tex
Add diagrams for normalize2 example
Shift normalizeInc example to a later section

\section{Introduction}

@@

However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take. The following function is similar to the one from before, except that it offers two unique \emph{but mutually exclusive} approaches to fusion:

\TODO{If the main contribution is fusion for *filters* then we need to use a motivating example with a filter}

\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A1) (B2)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}

For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@). The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@). For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

\TODO{This program has filters and other interesting stuff, but doesn't have ``multiple unique best solutions''}
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 us
  = let sum1 = fold   (+) 0 us      (A1) (S1) (I1)
        gts  = filter (>0)  us      (A1) (S2) (I1)
        sum2 = fold   (+) 0 gts     (A1) (S2) (I2)
        nor1 = map  (/sum1) us      (A2) (S3) (I3)
        nor2 = map  (/sum2) us      (A2) (S4) (I3)
    in (nor1, nor2)
\end{code}
Our approach, @A1-A2@, requires only two loops for this particular case.
Stream fusion (@S1-S4@) is unable to fuse @sum1@ and @gts@ together, as there is nothing to be inlined between them, so requires four whole traversals of the input array, @us@.
Similarly, existing imperative solutions (@I1-I3@) would be unable to fuse @gts@ and @sum2@ together, as the iteration size is completely different.

@@

\begin{itemize}
\item   
We use existing integer linear programming clustering formulations such as Megiddo\cite{megiddo1998optimal} and Darte\cite{darte2002contraction},
and modify them to express constraints on fusing loops of different sizes if they can be fused into generator loops of the same size.
\TODO{ref}
% We present an algorithm to cluster the operators of a data flow graph into sub-graphs that can be fused into individual imperative loops.
% We express the constraints on fusion as an integer linear program, and use an off-the-shelf solver to compute a valid clustering. \TODO{ref}

>>>>

%!TEX root = ../Main.tex
\section{Introduction}

@@

However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take. The following function is similar to the one from before, except that it offers two unique \emph{but mutually exclusive} approaches to fusion. A diagram of the data flow graph for this example is on the next page.

\TODO{This program has filters and other interesting stuff, but doesn't have ``multiple unique best solutions''}

\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 us
  = let sum1 = fold   (+) 0 us      (A1) (S1) (I1)
        gts  = filter (>0)  us      (A1) (S2) (I1)
        sum2 = fold   (+) 0 gts     (A1) (S2) (I2)
        nor1 = map  (/sum1) us      (A2) (S3) (I3)
        nor2 = map  (/sum2) us      (A2) (S4) (I3)
    in (nor1, nor2)
\end{code}
Our approach, @A1-A2@, requires only two loops for this particular case.
Stream fusion (@S1-S4@) is unable to fuse @sum1@ and @gts@ together, as there is nothing to be inlined between them, so requires four whole traversals of the input array, @us@.
Similarly, existing imperative solutions (@I1-I3@) would be unable to fuse @gts@ and @sum2@ together, as the iteration size is completely different.

@@

% NOTE: This set of bullets needs to fit on the first page, without spilling to the second.
\begin{itemize}
\item   
We use existing integer linear programming clustering formulations such as Megiddo\cite{megiddo1998optimal} and Darte\cite{darte2002contraction},
and modify them to express constraints on fusing loops of different sizes if they can be fused into generator loops of the same size.
\TODO{ref}
% We present an algorithm to cluster the operators of a data flow graph into sub-graphs that can be fused into individual imperative loops.
% We express the constraints on fusion as an integer linear program, and use an off-the-shelf solver to compute a valid clustering. \TODO{ref}

### 2014-04-07 2014betterfusionforfilters/section/04-ILP.tex
Add diagrams for normalize2 example
Shift normalizeInc example to a later section

\section{Integer linear program formulation}
While the different combinators are all implemented differently, certain aspects are shared between all; they operate on input arrays, use scalar variables in their worker functions, and produce output.
To simplify creation of the integer linear program formulation, we first convert a function to a dependency graph.
The dependency graph has a node for each combinator, and edges between two combinators when one uses another's output.
After the graph is created, it is converted to an integer linear program with an objective of the least memory traffic and loops, solved, and the solution converted to a clustering.

@@

\subsubsection{Satisfiability}
For any program $p$, there exists a trivial clustering with no fusion at all.
We can use this as the variable assignment of $ilp(p)$.
For each pair of nodes $m,n \in p$, $x_{mn} = 1$ --- no fusion is possible.
For the $\pi$ variables, we must find a topographical ordering of the nodes in $p$, which is simple since we are assured it is a dag.

>>>>

%!TEX root = ../Main.tex
\section{Integer linear program formulation}

\TODO{More justification of why we want to use a heavyweight technology like ILP to do the clustering. Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function.}
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A1) (B2)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}

For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@). The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@). For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

While the different combinators are all implemented differently, certain aspects are shared between all; they operate on input arrays, use scalar variables in their worker functions, and produce output.
To simplify creation of the integer linear program formulation, we first convert a function to a dependency graph.
The dependency graph has a node for each combinator, and edges between two combinators when one uses another's output.
After the graph is created, it is converted to an integer linear program with an objective of the least memory traffic and loops, solved, and the solution converted to a clustering.

@@

%!TEX root = ../Main.tex
\subsubsection{Satisfiability}
For any program $p$, there exists a trivial clustering with no fusion at all.
We can use this as the variable assignment of $ilp(p)$.
For each pair of nodes $m,n \in p$, $x_{mn} = 1$ --- no fusion is possible.
For the $\pi$ variables, we must find a topographical ordering of the nodes in $p$, which is simple since we are assured it is a dag.
